---
title: "codebook.Rmd"
author: "Milen Angelov"
date: "July 21, 2015"
output: 
  html_document: 
    keep_md: yes
---

## CodeBook
###Peer Assessment: Getting and Cleaning Data Course Project
This is code book that describes the variables, the data, and any transformations 
or work that you performed to clean up the data called CodeBook.md

Set default location of the data which will be used
```{r}
dataUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip";
```

If this script is run for the first time download and extract data
```{r}
if(!file.exists("./UCI HAR Dataset")){
    # Download source, on Mac OS X method should be specified
    download.file(dataUrl, destfile="./dataset.zip", method="curl");
    
    # Unzip it
    unzip("./dataset.zip");
}
```

Now define common functionality for reading, labeling and merging data from 
train and test datafiles. These function cover the following requirement:
1. Merges the training and the test sets for each set (activity, features and subject)
2. Appropriately labels the data set with descriptive variable names.

```{r}
## Function to read train and test subject data, label the column names and 
## return merged both dataset
getSubjectData <- function(){
    subject_train <- read.table("./UCI HAR Dataset/train/subject_train.txt",
                                   header = FALSE);
    
    subject_test <- read.table("./UCI HAR Dataset/test/subject_test.txt",
                                  header = FALSE);
    
    subject <- rbind(subject_train, subject_test);
    
    colnames(subject) <- c("Subject");
    
    subject;
}
```

now activity (and requirement 3 - Uses descriptive activity names to name the activities in the data set):
```{r}
## Function to read, name activities, merge and set column names of activity data.

getActivityData <- function(){
    
    activity_labels <- read.table("./UCI HAR Dataset/activity_labels.txt", 
                                  header = FALSE);
    
    y_train <- read.table("./UCI HAR Dataset/train/y_train.txt",
                         header = FALSE);
    
    y_test <- read.table("./UCI HAR Dataset/test/y_test.txt",
                          header = FALSE);
    
    activities <- rbind(y_train, y_test);
    
    #Uses descriptive activity names to name the activities in the data set
    activities$V1 <- factor(activities$V1, levels = activity_labels$V1);
    
    colnames(activities) <- c("Activity");
    
    activities;
}
```

and finally features:
```{r}
## Function to read, name columns and return merged both train and test 
## features data
getFeaturesData <- function(){
    X_train <- read.table("./UCI HAR Dataset/train/X_train.txt",
                          header = FALSE);
    X_test <- read.table("./UCI HAR Dataset/test/X_test.txt",
                         header = FALSE);
    
    featuresColNames <- read.table("./UCI HAR Dataset/features.txt");
    
    features <- rbind(X_train, X_test);
    colnames(features) <- featuresColNames$V2;
    
    features;
    
}
```

Now, lets merge features, subject and activity by columns:
```{r}
## Merges the training and the test sets to create one data set.
mergedData <- cbind(getFeaturesData(),
                    getSubjectData(),
                    getActivityData());
```
  
Let's check the structure and dimmension of the newly created dataset:
```{r}
str(mergedData);
dim(mergedData);
```

Extracts only the measurements on the mean and standard deviation for each measurement.

```{r}
## Get only the columns which include 'mean' or 'std' in their names
columnsWithMeasurement <- grep(".*mean.*|.*std.*", 
                               names(mergedData), 
                               ignore.case = TRUE);
```

Create a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r}
cleanedData <- mergedData[,columnsWithMeasurement];
```

Let's see how the new dataset looks like:
```{r}
str(cleanedData);
dim(cleanedData);
```

And create txt file to be uploaded containing dataset which was just created
```{r}
## File with the tidy data set created above to be uploaded as a txt file 
## created with write.table() using row.name=FALSE
write.table(cleanedData,
            file = "tidydataset.txt",
            sep = ",",
            row.names = FALSE);
```

